---
title: "Mini Data-Analysis Deliverable 1"
output: html_document
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1.  Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

<!-- -->

    install.packages("devtools")
    devtools::install_github("UBC-MDS/datateachr")

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
library(dplyr)
```

3.  Make a repository in the <https://github.com/stat545ubc-2022> Organization. You will be working with this repository for the entire data analysis project. You can either make it public, or make it private and add the TA's and Lucy as collaborators.

# Instructions

## For Both Milestones

-   Each milestone is worth 45 points. The number of points allocated to each task will be annotated within each deliverable. Tasks that are more challenging will often be allocated more points.

-   10 points will be allocated to the reproducibility, cleanliness, and coherence of the overall analysis. While the two milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as two chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

## For Milestone 1

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-1.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work below --->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to the mini-analysis GitHub repository you made earlier, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 45 points: 43 for your analysis, 1 point for having your Milestone 1 document knit error-free, and 1 point for tagging your release on Github.

# Learning Objectives

By the end of this milestone, you should:

-   Become familiar with your dataset of choosing
-   Select 4 questions that you would like to answer with your data
-   Generate a reproducible and clear report using R Markdown
-   Become familiar with manipulating and summarizing your data in tibbles using `dplyr`, with a research question in mind.

# Task 1: Choose your favorite dataset (10 points)

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

-   *apt_buildings*: Acquired courtesy of The City of Toronto's Open Data Portal. It currently has 3455 rows and 37 columns.

-   *building_permits*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 20680 rows and 14 columns.

-   *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

-   *flow_sample*: Acquired courtesy of The Government of Canada's Historical Hydrometric Database. It currently has 218 rows and 7 columns.

-   *parking_meters*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 10032 rows and 22 columns.

-   *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

-   *vancouver_trees*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

-   We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

-   Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

<!-------------------------- Start your work below ---------------------------->

1: CHOICE_1\_vancouver_trees\
2: CHOICE_2\_parking_meters\
3: CHOICE_3\_apt_buildings\
4: CHOICE_4\_building_permits\

<!----------------------------------------------------------------------------->

1.2 One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*.

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you below, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

<!-------------------------- Start your work below ---------------------------->

**1: CHOICE_1\_vancouver_trees\**
    -20 columns
    -146,611 rows
    -Data frame is a tibble
```{r}
#The dimensions are 20 columns and 146,611 rows
dim(vancouver_trees) 
#There are 20 columns and this shows each name
head(vancouver_trees)
#The data is in the form of a tibble data frame
class(vancouver_trees)
```

**2: CHOICE_2\_parking_meters\**
    -22 columns
    -10,032 rows
    -Data frame is a tibble
```{r}
#The dimensions are 22 columns and 10,032 rows
dim(parking_meters) 
#There are 22 columns and this shows each name
head(parking_meters)
#The data is in the form of a tibble data frame
class(parking_meters)
```

**3: CHOICE_3\_apt_buildings\**
    -37 columns
    -3,255 rows
    -Data frame is a tibble
```{r}
#The dimensions are 37 columns and 3,255 rows
dim(apt_buildings) 
#There are 37 columns and this shows each name
head(apt_buildings)
#The data is in the form of a tibble data frame
class(apt_buildings)
```

**4: CHOICE_4\_building_permits\**
    -14 columns
    -20,680 rows
    -Data frame is a tibble
```{r}
#The dimensions are 14 columns and 20,680 rows
dim(building_permits) 
#There are 37 columns and this shows each name
head(apt_buildings)
#The data is in the form of a tibble data frame
class(apt_buildings)
```

<!----------------------------------------------------------------------------->

1.3 Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 2. What lead you to choose these 2? Briefly explain your choices below, and feel free to include any code in your explanation.

<!-------------------------- Start your work below ---------------------------->
I will choose the vancouver_trees as well as parking_meters. These two data sets are both large (146,611 and 10,032 rows respectively) with 20 and 22 columns each. They are interesting data sets which will make them more rewarding to work with.
<!----------------------------------------------------------------------------->

1.4 Time for the final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with each dataset. Note them down below, and make your final choice based on what seems more interesting to you!

<!-------------------------- Start your work below ---------------------------->
Option 1: vancouver_trees: Does having a root barrier have a negative effect on tree diameter? (I will choose this one as it is the most relevant to my actual work in forestry/botany).
Option 2: parking_meters: What are the effects of geo_local_area on whether people pay to park with a credit card?
<!----------------------------------------------------------------------------->

# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. Remember, marks will be awarded for completion of the 4 tasks, but 10 points of the whole project are allocated to a reproducible and clean analysis. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset (15 points)

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE under Task 3**.

2.1 Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. Use *dplyr* and *ggplot*.

1.  Plot the distribution of a numeric variable.
2.  Create a new variable based on other variables in your data (only if it makes sense)
3.  Investigate how many missing values there are per variable. Can you find a way to plot this?
4.  Explore the relationship between 2 variables in a plot.
5.  Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
7.  Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8.  Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

<!-------------------------- Start your work below ---------------------------->
See work by each task
<!----------------------------------------------------------------------------->

# Task 3: Write your research questions (5 points)

So far, you have chosen a dataset and gotten familiar with it through exploring the data. Now it's time to figure out 4 research questions that you would like to answer with your data! Write the 4 questions and any additional comments at the end of this deliverable. These questions are not necessarily set in stone - TAs will review them and give you feedback; therefore, you may choose to pursue them as they are for the rest of the project, or make modifications!

<!--- *****START HERE***** --->
**Part 2: Exploring the data**
Here I will choose 4 of the tasks and complete them below with the data set vancouver_trees. I have chosen tasks 1,4,6,8 due to the nature of my data set. It is a large data set with a lot of categorical variables and a lot of variables that are potentially dependent on one another.

  1.  Plot the distribution of a numeric variable: I will plot the distribution of tree diameters

```{r}
#This is a ggplot showing the distribution of tree diameters in a bar graph to better visualize the main numerical column of the data.
ggplot(vancouver_trees, aes(x=diameter)) + geom_histogram(colour="black", fill="white")+xlim(0, 60)+ggtitle("Distribution of Tree Diameters")
```
 
  4.  Explore the relationship between 2 variables in a plot.
```{r}
#Here I will count how many of the entries  have a root barrier compared with those that do not to better understand what the data looks like.
count(vancouver_trees, root_barrier)
#Here I will plot  a stripplot that shows the relationship between diameter and whether there is a root barrier present or not which is directly related to one of my future questions. 
ggplot(vancouver_trees, aes(root_barrier, diameter)) + geom_point(alpha = 0.5, position="jitter")

#Here I will plot  a stripplot that shows the relationship between diameter and whether there is a curb present or not.
ggplot(vancouver_trees, aes(curb, diameter)) + geom_point(alpha = 0.5, position="jitter")
```
  
  6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
```{r}
#This makes a boxplot with a comparison of diameter between trees that have and do not have a root barrier. 
ggplot(vancouver_trees, aes(root_barrier, diameter)) + geom_boxplot()
```
  
  8.  Use a density plot to explore any of your variables (that are suitable for this type of plot).
```{r}
#This makes a histogram showing the density of the height range scales.
ggplot(vancouver_trees, aes(height_range_id)) +geom_histogram(aes(y = ..density..))
```
  
  
  **Task 3: Write your research questions**
  1. Does having a root barrier significantly decrease the average diameter of a tree?
  2. Is height correlated to diameter in all tree species?
  3. Does having a curb present next to the tree significantly decrease the average diameter of a tree?
  4. Does having a curb present next to the tree significantly decrease the average height range of a tree?

# Task 4: Process and summarize your data (13 points)

From Task 2, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions.

### 1.1 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

**Task 1.1**

Question 1. Does having a root barrier significantly decrease the average diameter of a tree?
  These tasks were each helpful to visualize the data. The boxplot is difficult to interpret as the boxes are so short/narrow. A statistical test would be more decisive for conclusions. The distribution of diameters is interesting to look at, but it does not answer the question and the same with the density plot. The boxplot would be the most effective for answering the questions, but again, it is difficult to interpret.
```{r}
# Make a boxplot relating the two variables to look at their relationship
ggplot(vancouver_trees, aes(root_barrier, diameter)) + geom_boxplot()
#show the distribution of diameters in a histogram
ggplot(vancouver_trees, aes(x=diameter)) + geom_histogram(colour="black", fill="white")+xlim(0, 60)+ggtitle("Distribution of Tree Diameters")

#This will make a density plot of diameters
ggplot(vancouver_trees, aes(diameter)) +geom_histogram(aes(y = ..density..))+xlim(0,60)

```

 Question 2. Is height correlated to diameter in all tree species?
  The first plot is a boxplot by height range id showing diameters. It is incredibly helpful for the question as it shows that as the height increases the diameter does as well. Then, I filtered the data to only include trees that are not on a curb and remade the same boxplot to see if this had a visible effect. The data looks fairly similar and statistical tests would be needed to determine significance.
```{r}
# Make a boxplot relating the two variables to look at their relationship
ggplot(vancouver_trees, aes(factor(height_range_id), diameter)) + geom_boxplot()

# Filter the data to make a set of only trees without a curb
van_trees_noCurb=vancouver_trees %>% filter(curb == "N")
#now this will remake the same boxplot but with the filtered data.
ggplot(van_trees_noCurb, aes(factor(height_range_id), diameter)) + geom_boxplot()

```
 
Question 3. Does having a curb present next to the tree significantly decrease the average diameter of a tree?
This once again is difficult to tell the answer to the question due to the narrow span of the boxplots. Statistical tests are needed to make conclusions. The filtering changes the graph a little, but not a huge amount.

```{r}
#This will make a boxplot showing the presence of a curb as it relates to diameter
ggplot(vancouver_trees, aes(curb, diameter)) + geom_boxplot()

#this will remove the trees that have a root barrier from the data and then replot the boxplot with the new data set.
van_trees_noBar=vancouver_trees %>% filter(root_barrier == "N")
ggplot(van_trees_noBar, aes(curb, diameter)) + geom_boxplot()
```

Question 4. Does having a curb present next to the tree significantly decrease the average height range of a tree?
The boxplot is quite definitive that there is no difference between the average height range of trees with and without curbs. 
```{r}
#this will make a boxplot with the relation between curb presence and height range.
ggplot(vancouver_trees, aes(curb, height_range_id)) + geom_boxplot()

#This will make a density plot of height range
ggplot(vancouver_trees, aes(height_range_id)) +geom_histogram(aes(y = ..density..))
```


**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
This is interesting for question 1 as it pertains to the diameters and gives a better picture of the distribution of values
```{r}
#here I will calculate the maximum diameter and minimum diameter for each height class and then calculate the range.
maxDbh=aggregate(diameter~height_range_id,data=vancouver_trees,max)
minDbh=aggregate(diameter~height_range_id,data=vancouver_trees,min)
rangeDbh=maxDbh-minDbh
rangeDbh
#this is the mean diameter of trees in each height class
aggregate(diameter~height_range_id,data=vancouver_trees,mean)
#this is the median diameter of trees in each height class
aggregate(diameter~height_range_id,data=vancouver_trees,median)
#this is the standard deviation of the diameter of trees in each height class
aggregate(diameter~height_range_id,data=vancouver_trees,sd)
```

2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
This is interesting for quesetions 3 and 4 as they both pertain to curbs.
```{r}
#This will count the numver of observations for yes and no to a curb
(summary_data <- vancouver_trees %>%
    group_by(curb) %>%
    summarise(Count = n())  )

```

3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
This could be interesting to include in the question numbers 1,2, and 3 as they pertain to diameter. It is common to group diameter into classes that represent the major life stages of a tree.
```{r}
#Here I create a new column called DiamRanges that represent the tree diameters of three groups that define the life stage that the trees fit within.
vancouver_trees_new <- within(vancouver_trees, {   
  DiamRanges <- NA # need to initialize variable
  DiamRanges[diameter < 100] <- "Sapling"
  DiamRanges[diameter >= 100 & diameter < 300] <- "Medium_growth"
  DiamRanges[diameter >= 300] <- "Old_growth"
   } )
vancouver_trees_new
```

4.  Based on two categorical variables, calculate two summary statistics of your choosing.
This is interesting in general. It is specific to question 3 and 4 which pertain to curbs.
```{r}
#This gives the rows as curb and the columns as root_barrier. The four summary statistics provided are the number of each combination.
table(vancouver_trees$curb, vancouver_trees$root_barrier)
```


**Graphing:**

5.  Create a graph out of summarized variables that has at least two geom layers.
This question is going off of the previous one. It produces a graph that pertains to curb and root barrier status.
```{r}
#This is a bar chart in ggplot with many layers to specify all of the details and names.
ggplot(vancouver_trees, aes(curb, ..count..)) + geom_bar(aes(fill = root_barrier), position = "dodge")+xlab("Curb")+ggtitle("Curbs and Root Barriers")+ylab("Count")+scale_fill_discrete(name = "Root Barrier", labels = c("No", "Yes"))
```

6.  Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
```{r}
#this is a graph of the log of diameter by height range id which is interesting for question 2.
ggplot(vancouver_trees, aes(x = factor(height_range_id), y = diameter, color=height_range_id)) + xlab("Height Range ID")+ylab("log of Diameter")+geom_point()+ scale_y_continuous(trans='log10')+scale_fill_discrete(name = "Height Range ID")+ggtitle("Log of Diameter by Height Range ID Class")
```

7.  Make a graph where it makes sense to customize the alpha transparency.
```{r}
#adding some transparency to the graph allows a clearer image of where each point falls.
ggplot(vancouver_trees, aes(x = height_range_id, y = diameter)) + xlab("Height Range ID")+ylab("Diameter")+geom_point(alpha = 0.5)
```

8.  Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.
Interesting for questions 1 and 4 due to the subject being height class.
```{r}
#The default is 30 bins
ggplot(vancouver_trees, aes(height_range_id)) +geom_histogram(aes(y = ..count..))
#This plot has bins=15
ggplot(vancouver_trees, aes(height_range_id)) +geom_histogram(bins=15,aes(y = ..count..))
#this plot has bins=8
ggplot(vancouver_trees, aes(height_range_id)) +geom_histogram(bins=8,aes(y = ..count..))

#It looks like 15 bins makes the cleanest looking histogram. More bins for a larger data set makes sense. Since we only have 10 height classes, too many bins makes it too hard to read and too few makes the pattern not distinct.
```


Make sure it's clear what research question you are doing each operation for!

### 1.2 (3 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!-------------------------- Start your work below ---------------------------->
Each question requires more details to be explored and requires statistical tests to clarify the results shown in the plots. It looks like **question 2** is the most interesting with a clearer correlation between height class and diameter. This should be further explored to examine species levels in relation to height and diameter to determine which species have a stronger correlation between the two variables. I believe question 2 is also the most refined as it has interesting and simple results that will be interesting to explore.



<!----------------------------------------------------------------------------->

### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.
